Understanding the Socio-Economic Determinants of Voter Behavior: An Analysis of the 2020 U.S. Presidential Election

### 2024-03-12 paper/us_election_2020.qmd
fix structure issue
added appendix

# Introduction
Four years ago, the United States once again was at the forefront of an important electoral contest between the Democratic and Republican parties in the 2020 presidential election. The Democratic and Republican parties are represented by Donald Trump and Joe Biden respectively. Donald Trump achieved a remarkable victory in the 2016 election against Hillary Clinton. On the other hand, Joe Biden, the former vice-president under Barack Obama's administration from 2008 to 2016, led the Democratic campaign. He aimed to regain control from the Republican party. In the end, Biden won the election by 51.3% majority votes (@cnn_2020), which Trump questioned its credibility. Due to the COVID-19 pandemic in 2020, the vote system changed to in-mail voting instead of in-person voting, which was why Trump believed that some votes were stolen (@trump_stolenvote). Therefore, it is important for us to analyze why people voted for Biden and Trump in 2020 election and how various demographic, economic, and social factors may influence individuals' voting decisions. Perhaps then we could see why Biden eventually won. We analyze the 2020 election data set from Cooperative Election Study (CES) [citation needed] and build a prediction model for individual voting preference. This allows us to discover trends and address potential political, economic, and societal situations that may cause the eventual winning for Biden.

@@

## Data Source
In this report, we use the 2020 Cooperative Election Study (CES) (@citeCes2020) as out primary dataset. This is a long-established annual political survey of US. CES contains information about how Americans view Congress and hold their representatives accountable in elections, how they vote and their electoral experiences, and how their behaviors and experiences vary with political geography and social context. In 2020, there were 61,000 American adults completed the survey.

@@

raw_ces2020[1:5, 1:5] |>
  kable()
```

@@

Firstly, for the variable "votereg" indicating whether a respondent is registered to vote, the measurement process involves relying on self-reported information from survey respondents. The US voter registration system is known to be inaccurate and inefficient.While efforts are made to ensure accuracy, there may be instances of misreporting or misunderstanding of voter registration status, leading to potential inaccuracies in the data. According to [@inaccurate], there are millions of voter registrations that are no longer valid or inaccurate. There may also be cases when individuals falsely believed they are registered to vote because they reached the voting age but did not actually register at their local office.

Regarding "CC20_410," representing the presidential candidate the respondent voted for, data collection relies on self-reporting, which is subject to recall bias and social desirability bias. While the data collection occurred before the notorious January 6, 2021 United States Capitol Attack that resulted in international criticism and increased the unpopularity of Trump in mainstream media, Trump had made many speeches and taken actions throughout his presidency that contributed to his polarizing image. Among many voters, voting for Trump became associated with characteristics such as being uneducated, backward-thinking, and anti-science. As a result, individuals who voted for him may have felt social pressure to conceal or lie about their support, potentially leading to underreporting of votes for Trump in the survey data.

@@

The "region" variable categorizes respondents based on their geographical location into four census regions: Northeast, Midwest, South, and West. However, regional boundaries can be arbitrary, and individuals living in bordering areas may have affiliations with multiple regions or identify more strongly with a neighboring region. For instance, states like Missouri and Kentucky are often considered part of the Midwest but also share cultural and economic ties with the South. This complexity can make individuals accurately reporting thier region challenging.

@@

The dataset is cleaned by rename the column names, specify the class of the columns, and change the numbers in the table to the corresponding description in the codebook to improve the readability. In addition, the cleaned data removes NA employment status since there are only 34 rows, which is a small amount of missing value compared with 61000 rows. After cleaning, 43534 rows of data and 4 variables remain.

@@

-------------------------------needs some explanation
```{r}
# We could use posterior predictive checks, introduced in Section 12.4, to show that the logistic regression approach is a good choice for this circumstance.
political_preferences1 <-
  readRDS(file = here::here("models/political_preferences1.rds"))

@@

```{r}
#| label: fig-race
#| fig-cap: The distribution of presidential preferences by race
#| echo: false
#| warning: false
#| message: false
ces2020 |>
  ggplot(aes(x = race, fill = voted_for)) +
  geom_bar(position = "dodge2") +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Race",
    y = "Number of respondents",
    fill = "Voted for"
  )
```

@@

@fig-region shows the relationship between employment and voting preferences. Across various employment statuses, there is a consistent trend of higher support for Biden compared to Trump. It can be observed that individuals holding full-time jobs show the most disparity in support between the two candidates, with Biden gaining substantially more support than Trump. The support for Biden and Trump among those with other employment statuses is relatively similar, with the support for Biden slightly higher than for Trump.
```{r}
#| label: fig-region
#| fig-cap: The distribution of presidential preferences by race
#| echo: false
#| warning: false
#| message: false
ces2020 |>
  ggplot(aes(x = employ, fill = voted_for)) +
  geom_bar(position = "dodge2") +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Employment status",
    y = "Number of respondents",
    fill = "Voted for"
  )
```

@@

# plot_ly(x=ces2020$race, y=ces2020$region, z=ces2020$employ, colors = c("red", "blue")) %>% 
#   add_markers(color = ces2020$voted_for, opacity = 0.03, size = 0.4) %>%
#   layout(xaxis = list(title = 'Race'), yaxis = list(title = 'Region'), zaxis = list(title = 'Employment Status')) %>%
#   layout(xaxis = list(titlefont = list(size = 22), tickfont = list(size = 22)),
#          yaxis = list(titlefont = list(size = 22), tickfont = list(size = 22)),
#          zaxis = list(titlefont = list(size = 22), tickfont = list(size = 22)))
```
<!-- This bubble plot (@fig-race_employ_region) illustrates four variables: race, region, employment status, and voting preference. Darker dots indicate a higher number of people, with red representing Trump supporters and blue for Biden supporters. Purple dots suggest similar support rates for both candidates, since blue and red makes purple in a 1:1 ratio. We can see that the darker dots are mostly in White, Black, Hispanic communities for full-time, part-time, or retired individuals. The darker dots are all blue, which indicates stronger support for Biden. It can also be noted that there are very few entirely red dots, which means that there is no significant disparities in voting percentages between Trump and Biden even among those groups which partially supported Trump. In those purple dots, we found that mostly are from people who are either Native American or from the South. -->

@@

Our results are summarized in @tbl-modelresults.

@@

modelsummary(
  list(
    "Support Biden" = political_preferences1
  ),
  statistic = "mad"
  )
```
Our results generally matches our justification. To avoid multicollinearity, the model excluded three variables from each category: race Asian, region Midwest, and employed full-time.The intercept represents the estimated log-odds of supporting Biden when all other predictors are held constant at their reference levels. In this case, the estimated log-odds of supporting Biden for individuals who are Asian, live in the Midwest, and are employed full-time is 0.806.

@@

Figure # shows range of coefficient estimates of our model within the 90% probability. Due to the fact that the credibility interval for race "Other" is particularly large, we cannot observe the trend of the 90% credibility intervals of other variables. Therefore we created figure # with the x axis limited from -10 to 10.

Combing figure # and figure #, we observe statistical significance for the coefficient estimates for students, the Northeast region, individuals identifying with two or more races, individuals identifying with "Other" racial category, Native Americans, Black individuals, and the intercept, Asians in the Midwest region who are employed full-time. The estimates are significant if the intervals do not cross 0. The value for the estimates are in log-odds, indicating that if the coefficient is positive, the individual supports Biden, if negative, the individual supports Trump.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-modelresults1
#| fig-cap: "Credible intervals for predictors of support for Biden 1"

modelplot(political_preferences1, conf_level = 0.9) +
  labs(x = "90% credibility interval")
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-modelresults2
#| fig-cap: "Credible intervals for predictors of support for Biden 2"
# Create the model plot
model_plot <- modelplot(political_preferences1, conf_level = 0.9)

# Modify the x-axis limits
model_plot + xlim(-5, 5) +  # Adjust the limits as needed
  labs(x = "90% Credibility Interval")
```

@@

In our analysis, respondents of different races show varying levels of support, with a majority of blacks, Hispanics, and Asians supporting Biden, and a majority of Middle Easterners and multiracials supporting Trump. The results for whites are more evenly distributed, with about the same percentage supporting Biden and Trump.

@@

2020 is the age of pandemics, as Trump's phrase "Chinese virus" led to a large number of Asians starting to support Biden. This is an extremely anti-Asian statement, although Covid was first found in China, but it does not mean that it is the Chinese virus. Trump's comments quickly festered in public opinion and led to a dramatic increase in hate incidents in the Asian American community (@Reja_2021). The repeated use of the "Chinese Virus" by a president of the country has contributed to an environment of racial discrimination and hatred against Asians, which has directly led to more Asians supporting Biden, whose policies are more open and democratic, and who values racial equality more.

Most white people vote relatively evenly, due to the fact that they are not affected by immigration or other policies. Whites hold a major voice and position in the US, most of the ideas and policies Trump proposes are favorable to whites. Therefore, he can gain support from most whites. However, there are some whites who want a democratic and free environment, and at the same time do not want to see racial discrimination and other phenomena, so they will choose to support Biden, who values equality.

@@

Biden's support base is concentrated in the Mid-Atlantic states, New England, the West Coast, Southwestern states, and certain regions around the Great Lakes. This support is particularly evident in notable cities within these regions, driven by various factors including liberal ideals and progressive policies. For instance, states like New York and Massachusetts support Biden due to their liberal-leaning populations, diverse demographics, and emphasis on social justices. Similarly, on the West Coast, states such as California and Washington are known for their progressive values and advocacy for issues like environmental conservation and LGBTQ+ rights, contributing to Biden's popularity in these areas. In the Southwest, states like Arizona and Nevada, with growing minority populations and concerns over immigration policies, also show significant support for Biden. Additionally, states around the Great Lakes, including Illinois and Michigan, align with Biden's platform due to their industrial heritage and strong labor unions.

@@

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
# 
# pp_check(first_model) +
#   theme_classic() +
#   theme(legend.position = "bottom")
# 
# posterior_vs_prior(first_model) +
#   theme_minimal() +
#   scale_color_brewer(palette = "Set1") +
#   theme(legend.position = "bottom") +
#   coord_flip()
```

>>>>

# Introduction
Four years ago, the United States once again was at the forefront of an important electoral contest between the Democratic and Republican parties in the 2020 presidential election. The Democratic and Republican parties are represented by Donald Trump and Joe Biden respectively. Donald Trump achieved a remarkable victory in the 2016 election against Hillary Clinton. On the other hand, Joe Biden, the former vice-president under Barack Obama's administration from 2008 to 2016, led the Democratic campaign. He aimed to regain control from the Republican party. In the end, Biden won the election by 51.3% majority votes (@cnn_2020), which Trump questioned its credibility. Due to the COVID-19 pandemic in 2020, the vote system changed to in-mail voting instead of in-person voting, which was why Trump believed that some votes were stolen (@trump_stolenvote). Therefore, it is important for us to analyze why people voted for Biden and Trump in 2020 election and how various demographic, economic, and social factors may influence individuals' voting decisions. Perhaps then we could see why Biden eventually won. We analyze the 2020 election data set from Cooperative Election Study (CES) (@citeCes2020) and build a prediction model for individual voting preference. This allows us to discover trends and address potential political, economic, and societal situations that may cause the eventual winning for Biden.

@@

## Data Source
In this report, we use the 2020 Cooperative Election Study (CES) (@citeCes2020) as our primary dataset. This is a long-established annual political survey of US. CES contains information about how Americans view Congress and hold their representatives accountable in elections, how they vote and their electoral experiences, and how their behaviors and experiences vary with political geography and social context. In 2020, there were 61,000 American adults completed the survey.

@@

# raw_ces2020[1:5, 1:5] |>
#   kable()
```

@@

Firstly, for the variable "votereg" indicating whether a respondent is registered to vote, the measurement process involves relying on self-reported information from survey respondents. The US voter registration system is known to be inaccurate and inefficient. While efforts are made to ensure accuracy, there may be instances of misreporting or misunderstanding of voter registration status, leading to potential inaccuracies in the data. According to [@inaccurate], there are millions of voter registrations that are no longer valid or inaccurate. There may also be cases when individuals falsely believed they are registered to vote because they reached the voting age but did not actually register at their local office.

Regarding "CC20_410," representing the presidential candidate the respondent voted for, data collection relies on self-reporting, which is subject to recall bias and social desirability bias. While the data collection occurred before the notorious January 6, 2021 United States Capitol Attack that resulted in international criticism and increased the unpopularity of Trump in mainstream media, Trump had made many speeches and taken actions throughout his presidency that contributed to his polarizing image. Among many voters, voting for Trump became associated with characteristics such as being uneducated, backward-thinking, and anti-science. As a result, individuals who voted for him may have felt social pressure to conceal or lie about their support, potentially leading to under-reporting of votes for Trump in the survey data.

@@

The "region" variable categorizes respondents based on their geographical location into four census regions: Northeast, Midwest, South, and West. However, regional boundaries can be arbitrary, and individuals living in bordering areas may have affiliations with multiple regions or identify more strongly with a neighboring region. For instance, states like Missouri and Kentucky are often considered part of the Midwest but also share cultural and economic ties with the South. This complexity can make individuals accurately reporting their region challenging.

@@

The dataset is cleaned by renaming the column names, specifying the class of the columns, and changing the numbers in the table to the corresponding description in the codebook to improve the readability. In addition, NA employment status is removed in the cleaned data file since there are only 34 entries, which is a small amount of missing value compared with 61000 entries. After cleaning, 43534 rows of data with 4 data features remain.

@@

## Model Implication
Figure # shows range of coefficient estimates of our model within the 90% probability. Due to the fact that the credibility interval for race "Other" is particularly large, we cannot observe the trend of the 90% credibility intervals of other variables. Therefore we created figure # with the x axis limited from -10 to 10.

Combining figure # and figure #, we observe statistical significance for the coefficient estimates for students, the Northeast region, individuals identifying with two or more races, individuals identifying with "Other" racial category, Native Americans, Black individuals, and the intercept, Asians in the Midwest region who are employed full-time. The estimates are significant if the intervals do not cross 0. The value for the estimates are in log-odds, indicating that if the coefficient is positive, the individual supports Biden, if negative, the individual supports Trump.

@@

```{r}
#| label: fig-race
#| fig-cap: The distribution of presidential preferences by race
#| echo: false
#| warning: false
#| message: false
ces2020 |>
  ggplot(aes(x = race, fill = voted_for)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Race",
    y = "Number of respondents",
    fill = "Voted for"
  )
```

@@

@fig-region shows the relationship between employment and voting preferences. Across various employment statuses, there is a consistent trend of higher support for Biden compared to Trump. It can be observed that individuals holding full-time jobs show the most disparity in support between the two candidates, with Biden gaining substantially more support than Trump. The support for Biden and Trump among those with other employment statuses is relatively similar, with the support for Biden slightly higher than for Trump.
```{r}
#| label: fig-region
#| fig-cap: The distribution of presidential preferences by race
#| echo: false
#| warning: false
#| message: false
ces2020 |>
  ggplot(aes(x = employ, fill = voted_for)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Employment status",
    y = "Number of respondents",
    fill = "Voted for"
  )
```

@@

# plot_ly(x=ces2020$race, y=ces2020$region, z=ces2020$employ, colors = c("red", "blue")) %>% 
#   add_markers(color = ces2020$voted_for, opacity = 0.03, size = 0.4) %>%
#   layout(xaxis = list(title = 'Race'), yaxis = list(title = 'Region'), zaxis = list(title = 'Employment Status')) %>%
#   layout(xaxis = list(titlefont = list(size = 22), tickfont = list(size = 22)),
#          yaxis = list(titlefont = list(size = 22), tickfont = list(size = 22)),
#          zaxis = list(titlefont = list(size = 22), tickfont = list(size = 22)))
```

@@

Our results are summarized in @tbl-modelresults1.

@@

modelsummary(
  list(
    "Support Biden" = political_preferences1
  ),
  statistic = "mad"
  )
```
Our results generally matches our expectation To avoid multicollinearity, the model excluded three variables from each category: race Asian, region Midwest, and employed full-time.The intercept represents the estimated log-odds of supporting Biden when all other predictors are held constant at their reference levels. In this case, the estimated log-odds of supporting Biden for individuals who are Asian, live in the Midwest, and are employed full-time is 0.806.

@@

@@

In our analysis, respondents of different races show varying levels of support, with a majority of blacks, Hispanics, and Asians supporting Biden, and a majority of Native Americans and multi-racials supporting Trump. The results for whites are more evenly distributed, with about the same percentage supporting Biden and Trump.

@@

2020 is the age of pandemics, as Trump's phrase "Chinese virus" led to a large number of Asians starting to support Biden. This is an extremely anti-Asian statement: although COVID-19 was first found in China, but it does not mean that virus was from China. Trump's comments quickly festered in public opinion and led to a dramatic increase in hate incidents in the Asian American community (@Reja_2021). The repeated use of the "Chinese Virus" by a president of the country has contributed to an environment of racial discrimination and hatred against Asians, which has directly led to more Asians supporting Biden, whose policies are more open and democratic, and who values racial equality more.

Most white people vote relatively evenly, due to the fact that they are not affected by immigration or other policies. Whites hold a major voice and position in the US, most of the ideas and policies Trump proposes are favorable to whites. Therefore, he can gain support from most whites. However, there are some whites who want a democratic and equal environment, and at the same time do not want to see racial discrimination and other phenomena, so they will choose to support Biden, who values equality.

@@

Biden's support base is concentrated in the Mid-Atlantic states, New England, the West Coast, Southwestern states, and certain regions around the Great Lakes. This support is particularly evident in cities within these regions, driven by various factors including liberal ideals and progressive policies. For instance, states like New York and Massachusetts support Biden due to their liberal-leaning populations, diverse demographics, and emphasis on social justices. Similarly, on the West Coast, states such as California and Washington are known for their progressive values and advocacy for issues like environmental conservation and LGBTQ+ rights, contributing to Biden's popularity in these areas. In the Southwest, states like Arizona and Nevada, with growing minority populations and concerns over immigration policies, also show significant support for Biden. Additionally, states around the Great Lakes, including Illinois and Michigan, align with Biden's platform due to their industrial heritage and strong labor unions.

@@

## Posterior predictive check
In @fig-post_dist, we implement a posterior distribution. This compares who people voted for in reality with the prediction results from the posterior distribution from our logistic regression model. It can be observed that the posterior distribution fits perfectly with the actual data, it suggests that the model accurately captures the observed data patterns. So the posterior is able to generate simulated data that closely resembles the actual data (@bayes_post). This is a positive sign because it indicates that our logistic regression model is a good representation of the actual voting preferences in the 2020 election data.

```{r}
#| label: fig-post_dist
#| fig-cap: Posterior distribution for logistic regression model 
#| echo: false
#| warning: false
#| message: false

@fig-post_prior compares the posterior with the prior. This compares the prior distribution of parameters with the posterior distribution of parameters in our logistic regression model. We can see that majority of the model parameters do not change even after data are taken into account. This shows that the observed data matches with our initial belief and expectation about the voting preferences for 2020 presidential election. However, for "raceOther", the posterior distribution shifts from its prior after we input observed data. This is likely suggesting that the observed data for "raceOther" strongly contradicts our initial belief. Yet, this is not a big problem since the percentage of the racial group "raceOther" is 0.86% (473/43534 from @tbl-summary) in our dataset.

```{r}
#| label: fig-post_prior
#| fig-cap: Comparing the posterior with the prior 
#| echo: false
#| warning: false
#| message: false

#### Read data ####
analysis_data <- read_csv(here::here("data/analysis_data/cleaned_ces2020.csv"))

# Convert variables to factors
analysis_data$race <- factor(analysis_data$race)
analysis_data$region <- factor(analysis_data$region)
analysis_data$employ <- factor(analysis_data$employ)

# Create voted_for variable in binary form
analysis_data$voted_for_binary <- ifelse(analysis_data$voted_for == "Biden", 1, 0)

# Model 1 for n = 1000
set.seed(123)

ces2020_reduced <- 
  analysis_data |> 
  slice_sample(n = 1000)

posterior_vs_prior(political_preferences1) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Credibility interval

```{r}
#| label: fig-conf
#| fig-cap: Comparing the posterior with the prior 
#| echo: false
#| warning: false
#| message: false
plot(
  political_preferences1,
  "areas"
)
```
```{r}
#| label: fig-conf_lim
#| fig-cap: Comparing the posterior with the prior 
#| echo: false
#| warning: false
#| message: false
#| 
model_plot <- plot(
  political_preferences1,
  "areas"
)
# Modify the x-axis limits
model_plot + xlim(-10, 10) +  # Adjust the limits as needed
  labs(x = "90% Credibility Interval")
```
## Credibility Interval

```{r}
# #| echo: false
# #| eval: true
# #| warning: false
# #| message: false
# #| label: fig-modelresults1
# #| fig-cap: "Credible intervals for predictors of support for Biden 1"
# 
# modelplot(political_preferences1, conf_level = 0.95, size = 0.1) +
#   labs(x = "90% credibility interval")
```

```{r}
# #| echo: false
# #| eval: true
# #| warning: false
# #| message: false
# #| label: fig-modelresults2
# #| fig-cap: "Credible intervals for predictors of support for Biden 2"
# # Create the model plot
# model_plot <- modelplot(political_preferences1, conf_level = 0.9)
# 
# # Modify the x-axis limits
# model_plot + xlim(-5, 5) +  # Adjust the limits as needed
#   labs(x = "90% Credibility Interval")
```
